Preliminary results

### 2019-03-07 reports/preliminary_results.Rmd
new version with correct comorbidity data. See (other project but related): https://www.pivotaltracker.com/story/show/164274282

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
load("../cache/df.RData")
library(tidyverse)
```

@@

I have used data from the "new linkage data base". Primary prosthesis inserted 2008-2015 were included. Note that we should have no problem with censoring here since death dates are recorded up until February 2018. Hence even if patients were operated during 2015, they are not automatically censored at the end of that year. (Censoring would only occour in the case of migration etc.)
Comorbidity indices based on Elixhauser van Walraven (ECI), Charlson's original index (CCI) and a modified Risk Rx IV index (modified by Anne Garland) were used.

@@

education = 
* low = pre-gymnasium
* middle = gymnasium
* high = post-gymnasium

@@

BMI and age are approximately normally distributed. CCI and ECI have very little data (most of the patients have no recorded comorbidities). It should be discussed whether diagnoses set during hospitalization should be included. There are some negative ECI values as well (due to the van Walraven index).

@@

Type of hospital, sex, ASA, civil status, education and prosthesis group looks as expected.

We have few patients with reversed hybrids (unbalanced sample).

@@

Proportions for most variables are roughly the same as in earlier versions, although not exact due to the extended time frame of the study period. 
Comorbidity data differ but this is currently under discussion.

@@

An alternative to down-sampling is up-sampling or a combination of the two [@Lunardon2014].
An additional benefit of down-sampling however is that we will also reduce the sample size (without losing too much power). This is beneficial since step-wise regression techniques around 100 variables and a large data set will otherwise be very time consuming.

@@

A draw-back with the down-sampling however is that we might get a sample that is not representative for the larger population. Our estimates might be very sensitive to the individuals who just happened to show up in the sample. If we perform a step-wise regression on two different samples, it would therefore be possible that each data set will provide two different set of models (based on different predictors) [@Lukacs2010; @Sauerbrei1999; @Zellner2004].

@@

Note that variables that are highly correlated could not be included in the same selection step, since it will then be arbitrary which one of those to select. I therefore used four independent step-wise regression procedures to each re-sample evaluating:

1. General variables including comorbidity indices
2. Individual ECI diagnosis
3. Individual CCI diagnosis
4. Individual Risk Rx IV diagnosis

Of course in the first group, it could be argued that the different comorbidity indices are correlated with each other. We did not see this in practice however.

@@

It is usually recommended to only include variables that are in fact chosen in around 90 %
 (or even 95 % of the cases). This would be age, ASA, medication for inflammatory pain and sex in our case. We could use a 75 % limit however since this is not our final model.

Note that heart infarction and renal disease did not show up here. Those were included last time but we can see from the table above that those were in fact choose in less than 75 % of our Bootstrap-samples. It might be argued that those were in fact important predictors in the original sample. The goal of prediction modelling however is not to evaluate the past but to predict the future!

@@

The candidate variables above where found from four different starting points. We can now argue that those variables should not be heavily associated with each other. It would have been different if for example heart infarction would have been important if classified both by ECI, CCI and Rx. We could then combine those factors into one (which was done for the previous version of the manuscript). This might also be necessary with a different threshold than 75 %. It might nevertheless be relevant to use step-wise regression once more based on the smaller subset of candidate variables.

@@

Hence, only CCI, age and ASA = 3 seems to be important.

@@

Given the results above, we might be tempted do drop the Rx-variables. It would be nice if those were not needed since acquiring this data in clinical practice might be cumbersome. Sex on the other hand is almost always available anyway. Hence, we'll better keep it. This will yield an alternative model for comparison:

```{r}
round(modsums$modsum[[2]]$coefmat.full, 2)
```

Figure 2 in the earlier version of the manuscript also included models with only ASA, CCI, ECI and Risk Rx. We will do too for comparison and can add a model  with age and sex only as well.

@@

We now 7 models that can be compared based on predictive power, thus by area under the (ROC) curve (AUC).
Let's take another 1,000 bootstrap replicates and compare observations to predictions for those. Note that we do not only use the list of predictors for each model, but also the estimated coefficients. We thus evaluate the models from above, rather than fitting new ones and evaluate again. We do this to decrease overfitting. This will give us 1,000 ROC curves for each model. We could plot all of those but the resulting figure would be quite messy (and time consuming to draw). I therefore fitted a locally estimated scatter-plot smoothing (LOESS) line as an average ROC-curve based on individual bootstrap estimates.

Overfitting will nevertheless occur since we are in fact still using the training data. We thus need to re-calculate AUC-values one final time for the evaluation data set, the 10 % of the original data that we have not looked at yet. We could apply Bootstrap replicates to the test data as well in order to estimate uncertainty/confidence intervals. Those would probably we very wide though and are avoided to save (computational) time.

@@

There seems to be no relevant difference between the full and simplified model. This is good, since we can then avoid the cumbersome data retrieval of prescription data if using the model in practice. A simpler model with age and sex only, is almost as good, especially applied to a previously unseen data set (which is what we should care about).

It is evident that ROC curves from the training data over-estimate the predictive power of each model. This is clear since ROC curves for the evaluation set deviates less from the horizontal line.

@@

We can conclude that the winning model have the same AUC as the chosen model from the previous version of the manuscript (0.8). We could also conclude however that this is an optimistic estimate since we do not reach that far using the test data. We can also conclude that the full model is in fact no better than the slightly simplified model or the model with age and sex only.

It might seem strange that both CCI and ECI actually underestimate the AUC. We can see above however that corresponding curves based in the evaluation data are very close to the diagonal with just a few exceptions. We might suspect that the larger training set included some extreme values for those indices that became influential for the model building, but which were not present in the evaluation set. We said in the beginning that a common split between training data and evaluation data is 3:1. We had 9:1 (which we knew were risky). We did so because the outcome is so rare and we therefore wanted to maximize the available data for estimation.

Lets have a look at the extremes of the Elixhauser index (CCI is less extreme but have a similar tendency). Most patientd have ECI within the range -4 to 17. Those are excluded here:

```{r}
load("../cache/ECI_extremes.RData")
ECI_extremes %>% 
  filter(!ECI_index_walraven %in% -4:17)
```

Hence, there is only one individual with ECI = -14, and she was assigned to the training set. She had obesity, drug abuse and depression, all associated with a long and happy life according to Van Walraven (Sic!).

@@

One way to avaoid exrteme observations for ECI/CCI would be to transform those variables by the Yeo-Johnson transformation. I tried this but it did not help. It might therefore be reasonable tom simply truncate the scale (or even ignore/exclude those extreme cases).

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
load("../cache/df.RData")
library(tidyverse)
options(digits = 2)
```

@@

I have used data from the "new linkage data base". Primary prosthesis inserted 2008-2015 were included. Note that we should have no problem with censoring here since death dates are recorded up until February 2018. Hence even if patients were operated during 2015, they are not automatically censored at the end of that year. (Censoring would only occour in the case of emigration etc.)
Comorbidity indices based on Elixhauser van Walraven (ECI), Charlson's original index (CCI) and a modified Risk Rx IV index (modified by Anne Garland) were used.

@@

education =

* low = pre-gymnasium
* middle = gymnasium
* high = post-gymnasium

@@

BMI and age are approximately normally distributed. There are some negative ECI values as well (due to the van Walraven index).

@@

Type of hospital, sex, ASA, civil status, education and looks as expected.

@@

Proportions for most variables are roughly the same as in earlier versions, although not exact due to the extended time frame of the study period.

@@

An alternative to down-sampling is up-sampling or a combination of the two [@Lunardon2014].
An additional benefit of down-sampling however is that we will also reduce the sample size (without losing too much power). This is beneficial since step-wise regression techniques with around 100 variables and a large data set will otherwise be very time consuming.

@@

A draw-back with the down-sampling however is that we might get a sample that is not representative for the larger population. Our estimates might be very sensitive to the individuals who just happened to end up in the sample. If we perform a step-wise regression on two different samples, it would therefore be possible that each data set will provide two different sets of models (based on different predictors) [@Lukacs2010; @Sauerbrei1999; @Zellner2004].

@@

Note that variables that are highly correlated could not be included in the same selection step, since it will then be arbitrary which one of those to select. I therefore used four independent step-wise regression procedures to each re-sample:

1. General variables
2. Individual ECI diagnosis
3. Individual CCI diagnosis
4. Individual Risk Rx IV diagnosis

@@

It is usually recommended to only include variables that are in fact chosen in around 90 %
 (or even 95 % of the cases). We could use a 75 % limit however since this is not our final model.

@@

The candidate variables above where found from four different starting points. It might therefore be relevant to use step-wise regression once more based on the smaller subset of candidate variables to account for possible dependencies.

@@

@@

We will compare our model to models with ASA, CCI, ECI and Risk Rx, as in the previous version of the manuscript. We can include a model with only age and sex as well.

@@

We now 6 models that can be compared based on predictive power, thus by area under the (ROC) curve (AUC).
Let's take another 1,000 bootstrap replicates and compare observations to predictions for those. Note that we do not only use the list of predictors for each model, but also the estimated coefficients. We thus evaluate the models from above, rather than fitting new ones and evaluate again. We do this to decrease overfitting. This will give us 1,000 ROC curves for each model. We could plot all of those but the resulting figure would be quite messy (and time consuming to draw). I therefore fitted a locally estimated scatter-plot smoothing (LOESS) line as an average ROC-curve based on individual bootstrap estimates.

Overfitting will nevertheless occur since we are in fact still using the training data. We thus need to re-calculate AUC-values one final time for the evaluation data set, the 10 % of the original data that we have not looked at yet. We could apply Bootstrap replicates to the test data as well in order to estimate uncertainty/confidence intervals. Those would probably we very wide though and are avoided to save (computational) time. We could do this as a later step before summarising the results however.

@@

It is evident that ROC curves from the training data over-estimate the predictive power of each model, as well as that the differnece between models get over-estimated.

@@

We can conclude that the winning model have the same AUC as the chosen model from the previous version of the manuscript (0.8). We could also conclude however that this is an optimistic estimate since we do not reach that far using the test data.

It might seem strange that both CCI and ECI actually underestimate the AUC. We can see above however that corresponding curves based on the evaluation data are very close to the diagonal with just a few exceptions. We might suspect that the larger training set included some extreme values for those indices that became influential for the model building, but which were not present in the evaluation set. We said in the beginning that a common split between training data and evaluation data is 3:1. We had 9:1 (which we knew were risky). We did so because the outcome is so rare and we therefore wanted to maximize the available data for estimation.

Lets have a look at the extremes of the Elixhauser index (CCI is less extreme but have a similar tendency). Most patientd have ECI within the range -7 to 21. Those are excluded here:

```{r}
load("../cache/ECI_extremes.RData")
ECI_extremes %>% 
  filter(!ECI_index_walraven %in% -7:21)
```

Hence, there is only one individual with ECI = -11, and she was assigned to the training set. She had obesity, drug abuse and depression, all associated with a long and happy life according to Van Walraven (Sic!).

@@

One way to avaoid extreme observations for ECI/CCI would be to transform those variables by the Yeo-Johnson transformation. I tried this but it did not help. It might therefore be reasonable to simply truncate the scale (or even ignore/exclude those extreme cases).

@@

# References
